---
title: "Data Analysis Assignment #2 (75 points total)"
author: "Brinker, Ashley"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1: (5 points) ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}
hist(mydata$RATIO, main = "Frequency of Abalone Ratio", xlab = "Ratio")

qqnorm(mydata$RATIO)
qqline(mydata$RATIO)

kurtosis(mydata$RATIO)
skewness(mydata$RATIO)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}

mydata$L_RATIO <- log10(mydata$RATIO)

hist(mydata$L_RATIO, main = "Frequency of L_Ratio", xlab = "L_Ratio")

qqnorm(mydata$L_RATIO)
qqline(mydata$L_RATIO)

kurtosis(mydata$L_RATIO)
skewness(mydata$L_RATIO)

boxplot(mydata$L_RATIO ~ mydata$CLASS, main = "Boxplot of L_Ratio by Class", xlab = "Class", 
        ylab = "L_Ratio")

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
bartlett.test(mydata$L_RATIO ~ mydata$CLASS)
bartlett.test(mydata$RATIO ~ mydata$CLASS)

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: L_Ratio exhibits better conformance to a normal distribution. By looking at our histograms we can see that L_Ratio appears to be more normal than Ratio, and this also seems to be the case with our QQ plots. By using our results from the Bartlett test, we see a p-value of 0.5267 > 0.05 for L_Ratio, meaning we fail to reject the null hypothesis that there is no difference in variance among our classes. When testing Ratio, we get a p-value of 0.0002531 < 0.05, meaning we reject the null hypothesis that there is no difference in variance among our classes. ***


***#### Section 2 (10 points) ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
a1<- aov(L_RATIO ~ CLASS + SEX + CLASS:SEX, data = mydata)
a2<- aov(L_RATIO ~ CLASS + SEX, data = mydata)

summary(a1)
summary(a2)

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: The summaries above suggest that the interaction term of class:sex is non-significant, but that the terms class and sex are significant when considered individually. ***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(a2)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: All of the comparisons are statistically significant except for A1 and A2. This implies that the L_Ratio of groups A1 and A2 is not significantly different. These results do suggest that male and female abalones could be combined into a single category adults. They do not have a significant difference for L_Ratio.***


***####  Section 3: (10 points) ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)


```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}
infant_sub<- subset(x = mydata, subset = (mydata$TYPE == "I"))
adult_sub<- subset(x = mydata, subset = (mydata$TYPE == "ADULT"))

par(mfrow=c(1,2))
hist(infant_sub$VOLUME, main = "Frequency of Infant Volume", xlab = "Volume", ylim = c(0,150), 
     xlim = c(0,1000), col = "red")
hist(adult_sub$VOLUME, main = "Frequency of Adult Volume", xlab = "Volume", ylim = c(0,150),
     xlim = c(0,1000), col = "blue")

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The results suggest that the volume of adults is normally larger than that of infants; however, there still could be difficulties in separating infants from adults. Particularly in the volume range of 100 to 300, where the total frequency of these volumes are very similar for adults and infants. ***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}
mydata$L_SHUCK <- log10(mydata$SHUCK)
mydata$L_VOLUME <- log10(mydata$VOLUME)

par(mfrow= c(2,2))
plot(mydata$VOLUME, mydata$SHUCK, col = mydata$CLASS, pch = 16, xlab = "VOLUME", ylab = "SHUCK")
legend(x = 900, y = 50, legend = levels(mydata$CLASS), col = c(1:5), pch=16, title = "CLASS")

plot(mydata$L_VOLUME, mydata$L_SHUCK, col = mydata$CLASS, pch = 16, xlab = "L_VOLUME", ylab = "L_SHUCK")
legend(x = 2.75, y = .5, legend = levels(mydata$CLASS), col = c(1:5), pch=16, title = "CLASS")

plot(mydata$VOLUME, mydata$SHUCK, col = mydata$TYPE, pch = 16, xlab = "VOLUME", ylab = "SHUCK")
legend(x = 900, y = 50, legend = levels(mydata$TYPE), col = c(1:2), pch=16, title = "TYPE")

plot(mydata$L_VOLUME, mydata$L_SHUCK, col = mydata$TYPE, pch = 16, xlab = "L_VOLUME", ylab = "L_SHUCK")
legend(x = 2.75, y = .5, legend = levels(mydata$TYPE), col = c(1:2), pch=16, title = "TYPE")

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: The log-transformation seems to reduce variability. As volume and shuck increase, we see the older classes appear more frequently, similarly with younger classes appearing more with lower volume and shuck. Additionally, we notice that infant appears most often among lower volume and shuck, while adult appears most often among higher volume and shuck.  ***


***####   Section 4: (5 points) ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}
model<- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data = mydata)
summary(model)

```

**Essay Question:  Interpret the trend in CLASS level coefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: The trend in class level coefficient estimates seem to be decreasing as class level increases. ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: Type does not seem to be an important predictor in this regression, as the coefficient is significantly smaller than the other variables being addressed. ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: (5 points) ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}
hist(model$residuals, main = "Frequency of Regression Residuals", xlab = "Residuals")

qqnorm(model$residuals)
qqline(model$residuals)

kurtosis(model$residuals)
skewness(model$residuals)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}
par(mfrow = c(2,2))

plot(mydata$L_VOLUME, model$residuals, col = mydata$CLASS, pch = 16, xlab = "VOLUME", ylab = "Residuals")
legend(x = .5, y = -.15, legend = levels(mydata$CLASS), col = c(1:5), pch=16, title = "CLASS")

plot(mydata$L_VOLUME, model$residuals, col = mydata$TYPE, pch = 16, xlab = "VOLUME", ylab = "Residuals")
legend(x = .5, y = -.2, legend = levels(mydata$TYPE), col = c(1:2), pch=16, title = "CLASS")

boxplot(model$residuals ~ mydata$CLASS, xlab = "CLASS", ylab = "Residuals")
boxplot(model$residuals ~ mydata$TYPE, xlab = "TYPE", ylab = "Residuals")

bartlett.test(model$residuals ~ mydata$CLASS)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: The output suggests that volume could be useful in harvesting decisions. It often displays that abalones in classes A4 and A5 (those ready for harvesting) tend to have higher volumes. By using our results from the Bartlett test, we see a p-value of 0.4498 > 0.05, meaning we fail to reject the null hypothesis that there is no difference in variance of residuals among our classes. ***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: (5 points) ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383

# Proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /
  total.adults  # [1] 0.2476573

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /
  total.infants

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:
(med_inf_vol <- median(mydata$VOLUME[mydata$TYPE == "I"])) 

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) /
  total.infants

# Add code to calculate the proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) /
  total.adults

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.


# Add code to determine the median adult volume:
(med_adlt_vol <- median(mydata$VOLUME[mydata$TYPE == "ADULT"])) 

# Add code to calculate the proportion of infants harvested
sum(mydata$VOLUME[mydata$TYPE == "I"] > med_adlt_vol) /
  total.infants

# Add code to calculate the proportion of adults harvested
sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_adlt_vol) /
  total.adults

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}
plot(volume.value, prop.infants, type = "l", lwd = 5, col = "red")
lines(volume.value, prop.adults, lwd = 5, col = "blue")
legend(x = 800, y = .2, legend = levels(mydata$TYPE), col = c("blue", "red"), lty = 1, lwd = 5)
abline(h = .5, lty = 2)
abline(v = max_inf_vol)
abline(v = med_inf_vol)
abline(v = med_adlt_vol)
text(x = 170, y = .45, labels = "Median Infant: \n133.8214")
text(x = 420, y = .45, labels = "Median Adult: \n384.5584")
text(x = 560, y = .95, labels = "Largest Infant: \n526.6383")

```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The two median values allow us to visually see the cutoffs being considered for harvest. This allows us to compare which infants may be harvested preemptively, and which adults may not have been harvested at their maturity. ***

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

***#### Section 7: (10 points)  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}
difference<- ((1 - prop.adults) - (1 - prop.infants))

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}
par(mfrow = c(1,2))

plot(volume.value, difference, type = "l", lwd = 5)
lines(volume.value, smooth.difference, lwd = 5, lty = 2, col = "red")
abline(v = volume.value[which.max(smooth.difference)], lwd = 3)
text(x = 270, y = .4, labels = "Volume = 262.143", srt = 90)

plot(volume.value, prop.infants, type = "l", lwd = 5, col = "red")
lines(volume.value, prop.adults, lwd = 5, col = "blue")
legend(x = 800, y = .2, legend = levels(mydata$TYPE), col = c("blue", "red"), lty = 1, lwd = 5)
abline(h = .5, lty = 2)
abline(v = max_inf_vol)
abline(v = med_inf_vol)
abline(v = med_adlt_vol)
abline(v = volume.value[which.max(smooth.difference)], lwd = 3)
text(x = 190, y = .45, labels = "Med Infant: \n133.8214")
text(x = 440, y = .45, labels = "Med Adult: \n384.5584")
text(x = 585, y = .95, labels = "Largest Infant: \n526.6383")
text(x = 300, y = .8, labels = "Max Diff \n262.143")


```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the vlume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
# OR,
sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8: (10 points)  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

prot_inf_vol<- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786

sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > prot_inf_vol) /
  total.adults

sum(mydata$VOLUME[mydata$TYPE == "I"] > prot_inf_vol) /
  total.infants

```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}
plot(volume.value, prop.infants, type = "l", lwd = 5, col = "red")
lines(volume.value, prop.adults, lwd = 5, col = "blue")
legend(x = 800, y = .2, legend = levels(mydata$TYPE), col = c("blue", "red"), lty = 1, lwd = 5)
abline(h = .5, lty = 2)
abline(v = max_inf_vol)
abline(v = med_inf_vol)
abline(v = med_adlt_vol)
abline(v = volume.value[which.max(smooth.difference)], lwd = 3)
abline(v = prot_inf_vol)
text(x = 160, y = .45, labels = "Med Infant: \n133.8214")
text(x = 410, y = .45, labels = "Med Adult: \n384.5584")
text(x = 560, y = .95, labels = "Largest Infant: \n526.6383")
text(x = 280, y = .8, labels = "Max Diff \n262.143")
text(x = 235, y = .7, labels = "Zero A1 inf: \n206.786")

```


***#### Section 9: (5 points) ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}
plot(x = (1 - prop.infants), y = (1 - prop.adults), type = "l", col = "blue", lwd = 5)
abline(a = 0, b = 1, lty = 2, col = "red", lwd = 5)
points(x = 0.4982699, y = 0.9330656, col = "black", pch = 1, cex = 3, lwd = 5) #median infant
points(x = 0.02422145, y = 0.49933079, col = "black", pch = 1, cex = 3, lwd = 5) #median adult
points(x = 0.2871972, y = 0.8259705, col = "black", pch = 1, cex = 3, lwd = 5) #zero A1 infants
points(x = 0, y = 0.2476573, col = "black", pch = 1, cex = 3, lwd = 5) #protect all infants
points(x = 0.1764706, y = 0.7416332, col = "black", pch = 1, cex = 3, lwd = 5) #max diff

text(x = .5, y = .89, labels = "Med Infant: \n133.8214")
text(x = .06, y = .45, labels = "Med Adult: \n384.5584")
text(x = .25, y = .85, labels = "Zero A1 inf: \n206.786")
text(x = .06, y = .25, labels = "Protect All Infants: \n526.6383")
text(x = .2, y = .7, labels = "Max Diff \n262.143")



```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
auc(x = (1 - prop.infants), y = (1 - prop.adults))

```


***#### Section 10: (10 points) ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	
med_inf_harvest <- sum(mydata$VOLUME > med_inf_vol) / length(mydata$VOLUME)
prot_inf_harvest <- sum(mydata$VOLUME > max_inf_vol) / length(mydata$VOLUME)
med_adlt_harvest <- sum(mydata$VOLUME > med_adlt_vol) / length(mydata$VOLUME)
max_dif_harvest <- sum(mydata$VOLUME > volume.value[which.max(smooth.difference)]) / length(mydata$VOLUME)
zero_a1_harvest <- sum(mydata$VOLUME > prot_inf_vol) / length(mydata$VOLUME)

my_table<- matrix(c(max_inf_vol, 0.2476573, 0, prot_inf_harvest,
  med_inf_vol, 0.9330656, 0.4982699, med_inf_harvest,
  med_adlt_vol, 0.49933079, 0.02422145, med_adlt_harvest,
  volume.value[which.max(smooth.difference)], 0.7416332, 0.1764706, max_dif_harvest,
  prot_inf_vol, 0.8259705, 0.2871972, zero_a1_harvest), ncol = 4, byrow = TRUE)

colnames(my_table)<- c("Volume", "TPR", "FPR", "totalHarvest")
rownames(my_table)<- c("Protect All Infants", "Median Infants", "Median Adults", "Max Difference", 
                       "Zero A1 Infants")
my_table<- as.table(my_table)

my_table
```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: Since we want to harvest as many abalones as possible, while reducing the number of infant abalones harvested, we want to see a higher proportion in TPR and totalHarvest, while keeping a fairly low proportion in FPR. Protect all infants has the benfit of zero infants being harvested, but only harvests 17% of the total population and 49% of adult abalones. Median infants has a very high total harvest and adult harvest rate, at 81% and 93% respectively. However, this method harvests nearly 50% of infant abalones, which could lead to population issues down the line. Median adults has nearly a 50% harvest rate for all adults, while keeping a very low harvest rate of 2% for infants. Max difference allows for a large proportion of adults to be harvested at 74%, and maintains a fairly low harvest rate of 18% for infants. Zero A1 infants has similar results with slightly higher values across the board. Depending on the ideal percentage of infant abalones needed to successfully continue the population, either using the max difference volume, or the zero A1 infants volume seem to be the most successful while not harming the infant population too drastically. ***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: 1. I would outline the various choices and tradeoffs, as the population of abalones may change over time, requiring a different volume cut off to help increase or decrease the proportion of infant abalones left after harvesting. 2. I would present the proportions displayed in the final table, as these values give insight to both adult, infant, and total populations. 3. I would suggest the max difference as the volume cut off, as this seems to be the best compromise between harvesting a fairly large percentage of adult abalones, while only harvesting a fairly small percentage of infant abalones. 4. I would suggest for future abalone studies of this type that we address the best harvest time by age, as we had to reclassify some infants as adults, based on their class. ***