---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Brinker, Ashley"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)

table(mydata$CLASS, mydata$RINGS)

```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: The variables length, diam, height, whole, shuck, rings, volume and ratio are ratio level variables, as they are all metric data that has the possibility of an absolute zero. The variable sex is a nominal level variable, as it is a non-metric variable used solely for classification. The variable class is an ordinal level variable as it categorizes based on the rings (which are based on age), meaning these categories can be related to each other (i.e. A1 is a group of abalones collectively younger than those in group A2). We can use Pearson's second coefficient of skewness (3*(mean-median)/std dev) to evaluate our ratio data where we find all variables except rings fall within -.5 and .5 skewness and are approximately symmetric. Rings is moderately skewed to the right with a skewness between .5 and 1. This can be confirmed by looking at the respective histograms.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
mytable<- table(mydata$SEX, mydata$CLASS)
addmargins(mytable)

barplot(mytable, main = "Distribution of Abalone by Class and Sex", xlab = "Class: A1-A5", ylab = "Frequency", legend = TRUE, ylim = c(0,400))

```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: The distribution of sex seems to show that abalones develop reproductive organs over the course of their life, as the proportion of infant abalones per class decreases as class (which is fundamentally based on age) increases. Alternately, this could address difficulties associated with determining sex. Additionally, we can note that some abalones in the A4 and A5 class are still infants; therefore, we could conclude that some abalones do not develop reproductive organs during the timeline measured, or they were unable to be categorized by researchers. The distribution by class is alarming because you would expect the youngest class to contain the highest frequency; however, is has the lowest. This could be due to the fact that class is firstly determined by the variable rings, which does not apply to abalones under the age of 1.5 years (as having 0 rings would mean the abalone is 1.5 years old). Since our minimum value for rings is 3, the classes here address abalones 4.5 years or older.  ***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work<-mydata[sample(x = c(1:1036), size = 200, replace=FALSE),]

plot(work[,2:6])

```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
plot(mydata$WHOLE, mydata$VOLUME, col = mydata$CLASS, main = "Abalones: Whole Weight vs Volume", xlab = "Whole Weight (g)", ylab = "Volume (cm^3)", pch=16)
legend(1, 950, legend = levels(mydata$CLASS), col = c(1:5), pch=16, title = "Class")

```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
plot(mydata$WHOLE, mydata$SHUCK, col = mydata$CLASS, main = "Abalones: Whole Weight vs Shuck Weight", xlab = "Whole Weight (g)", ylab = "Shuck Weight (g)", pch=16)
abline(a = 0, b = max(mydata$RATIO))
legend(1, 150, legend = levels(mydata$CLASS), col = c(1:5), pch=16, title = "Class")
text(280, 75, labels = "Max Ratio of \nShuck to Whole", srt = 17)

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: The variability of both graphs is similar. Both displays follow the logic that larger abalones by whole weight tend to have a larger volume or larger shuck weight, respectively. Additionally the classes, which relate to age, tend to have larger sizes relating to older abalones. ***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
#create subsets by sex
ratio_female<- subset(mydata, SEX == "F", select = RATIO)
ratio_male<- subset(mydata, SEX == "M", select = RATIO)
ratio_infant<- subset(mydata, SEX == "I", select = RATIO)

#histograms
par(mfrow = c(3,3))
hist(ratio_female$RATIO, main = "Ratio: Female Abalones", breaks =seq(from=0, to=.4, by=.05), 
     xlab = "Ratio", col = "red", ylim = c(0, 200))
hist(ratio_male$RATIO, main = "Ratio: Male Abalones", breaks =seq(from=0, to=.4, by=.05), 
     xlab = "Ratio", col = "blue", ylim = c(0, 200))
hist(ratio_infant$RATIO, main = "Ratio: Infant Abalones", breaks =seq(from=0, to=.4, by=.05), 
     xlab = "Ratio", col = "green", ylim = c(0, 200))

#boxplots
boxplot(ratio_female$RATIO, main = "Ratio: Female Abalones", ylim = c(0, .35), col = "red", 
        horizontal = TRUE)
boxplot(ratio_male$RATIO, main = "Ratio: Male Abalones", ylim = c(0, .35), col = "blue", 
        horizontal = TRUE)
boxplot(ratio_infant$RATIO, main = "Ratio: Infant Abalones", ylim = c(0, .35), col = "green", 
        horizontal = TRUE)

#QQ plots
qqnorm(ratio_female$RATIO, main = "Ratio: Female Abalones", col = "red")
qqline(ratio_female$RATIO)
qqnorm(ratio_male$RATIO, main = "Ratio: Male Abalones", col = "blue")
qqline(ratio_male$RATIO)
qqnorm(ratio_infant$RATIO, main = "Ratio: Infant Abalones", col = "green")
qqline(ratio_infant$RATIO)

```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: Looking at the QQ plots, we can observe the distance to the diagonal lines plotted in black. It appears that all three sexes are fairly normally distributed with infants having the least normality, regarding ratio. ***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
female_stats<- boxplot.stats(ratio_female$RATIO)
male_stats<- boxplot.stats(ratio_male$RATIO)
infant_stats<- boxplot.stats(ratio_infant$RATIO)

# FEMALE OUTLIERS
# 0.31176204 0.21216140 0.21465603 0.21306058 0.23497668 0.06733877
# MALE OUTLIERS
# 0.2609861 0.2378764 0.2345924 0.2356349 0.2286735
# INFANT OUTLIERS
# 0.2693371 0.2218308 0.2403394 0.2263294 0.2249577 0.2300704 0.2290478 0.2232339
# total outliers: 19

#compare ratio values to extreme lower and upper whisker values from boxplot.stats
outliers<- subset(mydata, (SEX == "F" & RATIO < female_stats$stats[1]) | 
                    (SEX == "F" & RATIO > female_stats$stats[5]) |
                    (SEX == "M" & RATIO < male_stats$stats[1]) |
                    (SEX == "M" & RATIO > male_stats$stats[5]) |
                    (SEX == "I" & RATIO < infant_stats$stats[1]) |
                    (SEX == "I" & RATIO > infant_stats$stats[5]))

kable(outliers)

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: There were a total of 19 outliers, 8 that were infants, 6 that were female, and 5 that were male. The extreme lower and upper whisker values were nearly identical for each sex; however, they were still handled individually to account for the small differences. Additionally, there are no outliers that belong to the class A4 in regards to the variable ratio. ***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
par(mfrow=c(2,2))

boxplot(mydata$VOLUME ~ mydata$CLASS, horizontal = TRUE, ylab = "Class", xlab = "Volume (cm^3)",
        col = "orange", main = "Volume vs. Class")
boxplot(mydata$WHOLE ~ mydata$CLASS, horizontal = TRUE, ylab = "Class", xlab = "Whole (g)",
        col = "purple", main = "Whole Weight vs. Class")

plot(mydata$VOLUME, mydata$RINGS, ylab = "Rings", xlab = "Volume (cm^3)", col = "orange", 
     main = "Volume vs. Rings")
plot(mydata$WHOLE, mydata$RINGS, ylab = "Rings", xlab = "Whole (g)", col = "purple",
     main = "Whole Weight vs. Rings")

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: I think these variables would perform well as predictors of age for classes A1 and A2. Starting around A3 (approximately 9 rings) you'll notice the scatterplots begin to level off, and the boxplots become more vertically stacked. Therefore, for classes A3-A5 these variables would not be a good predictor of age. ***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
agg_volume<- aggregate(mydata$VOLUME, by = list(mydata$SEX, mydata$CLASS), FUN = mean,
                       na.rm=TRUE)
agg_shuck<- aggregate(mydata$SHUCK, by = list(mydata$SEX, mydata$CLASS), FUN = mean,
                       na.rm=TRUE)
agg_ratio<- aggregate(mydata$RATIO, by = list(mydata$SEX, mydata$CLASS), FUN = mean,
                       na.rm=TRUE)

mymatrix<- matrix(agg_volume$x, nrow = 3)
dimnames(mymatrix)<- list(levels(mydata$SEX), levels(mydata$CLASS))

kable(mymatrix)

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
par(mfrow = c(2,2))
interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX, response = mydata$RATIO,
                 fun = mean, type = "l", ylim = c(.1, .18), legend = TRUE, trace.label = "Sex",
                 xlab = "Class", ylab = "Ratio", main = "Ratio Vs. Class", 
                 col = c("blue4", "red4", "green4"), lty = 1, lwd = 3, leg.bty = "o", 
                 fixed = TRUE)

interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX, response = mydata$VOLUME,
                 fun = mean, type = "l", ylim = c(0, 550), legend = TRUE, trace.label = "Sex",
                 xlab = "Class", ylab = "Volume (cm^3)", main = "Volume Vs. Class", 
                 col = c("blue4", "red4", "green4"), lty = 1, lwd = 3, leg.bty = "o", 
                 fixed = TRUE)

interaction.plot(x.factor = mydata$CLASS, trace.factor = mydata$SEX, response = mydata$SHUCK,
                 fun = mean, type = "l", ylim = c(0, 80), legend = TRUE, trace.label = "Sex",
                 xlab = "Class", ylab = "Shuck (g)", main = "Shuck Vs. Class", 
                 col = c("blue4", "red4", "green4"), lty = 1, lwd = 3, leg.bty = "o", 
                 fixed = TRUE)
```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: These plots raise the question of "do female abalones tend to have larger volume and larger shuck weight than male or infant abalones?" Additionally, we can ask "why does shuck weight tend to decline sharper at class A4 than volume at class A4?" This is also noticed in the steady decline of ratio for all three sexes. ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
par(mfrow=c(2,2))
infant_sub<- subset(mydata, SEX == "I" & RINGS < 16)
adult_sub<- subset(mydata, (SEX == "F" & RINGS < 16 ) | (SEX == "M" & RINGS < 16))

#boxplots for volume vs rings
boxplot(infant_sub$VOLUME ~ infant_sub$RINGS, horizontal = TRUE, ylim = c(0, 1100), 
        ylab = "Rings", xlab = "Volume (cm^3)", col = "orange", 
        main = "Infants: Volume by Rings")
boxplot(adult_sub$VOLUME ~ adult_sub$RINGS, horizontal = TRUE, ylim = c(0, 1100),
        ylab = "Rings", xlab = "Volume (cm^3)", col = "purple",
        main = "Adults: Volume by Rings")

#boxplots for whole vs rings
boxplot(infant_sub$WHOLE ~ infant_sub$RINGS, horizontal = TRUE, ylim = c(0, 400),
        ylab = "Rings", xlab = "Whole (g)", col = "orange",
        main = "Infants: Whole Weight by Rings")
boxplot(adult_sub$WHOLE ~ adult_sub$RINGS, horizontal = TRUE, ylim = c(0, 400),
        ylab = "Rings", xlab = "Whole (g)", col = "purple",
        main = "Adults: Whole Weight by Rings")

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: These displays suggest that growth slows after abalones have approximately 10 rings (associated with 11.5 years of age). There is a lot more variability in size among the adults when compared to the infants. ***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: Physical measurements are only accurate for age prediction up to approximately age 11.5 years (10 rings). After that, growth appears to slow/stagnate, and determining age (if desired past this maturity) is not likely. Additionally, with the issue of sexing the abalones, size does not seem to help significantly differentiate between Male and Female. The study background also mentioned that ring clarity can be an issue, which could be reflected in the data. ***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: I would ask how the sample was selected, how large the sample was, the primary goal of the study, if the population is finite or not, who the researchers are behind the study and the financial backing or sponsorship of the study. ***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: Observational studies risk bias among sample selection, error in measurements, and confounding variables and their impact. Because of these factors, observational studies cannot be used to represent causality. ***
